Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-19 tex/classical.tex
Rewording: up to p. 4 of sec. 2.3

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the functional \cite{bib:tolman,bib:huang}
%
\begin{equation}\label{eq:hfunctional}
   H=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Here, $f(\vec{v},t)$ is a homogeneous distribution function, \textit{i.e.}
$f$ is assumed to be independent of $\vec r$.
The functional $H$, originally introduced by Boltzmann in 1872,
describes a diluted gas, at temperature $T$, with a total energy $E$, a total
number of free classical particles $N$, and total volume $V$.

@@

The distribution functions, $\{f_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. In terms of $f_M(\vec{v},t)$ and the energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}
\begin{eqnarray}
    H_M' & = &  \int f_M(\vec{v},t) \ln f_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{Hcell},\\
    \mathcal{N}_M' & = & \int f_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\\
    \mathcal{E}_M' & = & \int f_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v.
\end{eqnarray}
\end{subequations}
%
Notice that assuming local homogeneous distribution functions means that we are surmising the 
local equilibrium hypothesis. In addition, $f_{M}(\vec{v},t)$ must satisfy the following
microcanonical restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

@@

We can see that the functional (\ref{eq:CH2}) contains a non-homogeneous
distribution function but keeps the same form of the Boltzmann's $H$ functional.

Inspired by the Hamilton principle, $H'$ will be maximized with those
restrictions to obtain the equilibrium distribution function for a classical gas
using the Lagrange multipliers method obtaining
%
\begin{eqnarray}
    \frac{\delta H'}{\delta f_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f_J(\vec{v}')}\left[
        f_M(\vec{v})\ln f_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f_M(\vec{v})}{\delta f_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f_M(\vec{v})}{\delta f_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Isolating the distribution function 
%
\begin{eqnarray}
    \ln f_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{relacion1},
\end{eqnarray}
%
where $C$ is a constant. It is trivial that using the variational procedures we
obtain the form of the Maxwell-Boltzmann distribution.

@@

We can approximate the logarithm function in (\ref{dH1-1}) to its first-order
Taylor series around $\bar f_n(\vec{v}) g_{nM}(\vec{v},t)=0$ 
%
\begin{equation}\label{lnapproximationclassical}
    \ln [\bar{f}(\vec{v})+\bar{f}(\vec{v}) g_{M}(\vec{v},t)] \approx
    \ln [\bar{f}(\vec{v})]+ g_{M}(\vec{v},t).
\end{equation}
%
With the previous approximation, (\ref{dH1-1}) obtains the following form 
%
\begin{eqnarray}
    \frac{dH'}{dt}&=&\sum_{M=1}^{K} \int \bar f(\vec{v})\left[
      1+\ln \bar f(\vec{v})+g_M(\vec{v},t)
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v,
\end{eqnarray}
%
using (\ref{relacion1}) with homogeneous multipliers
%
\begin{equation}\label{dH1-2}
    \frac{dH'}{dt} = \sum_{M=1}^{K}\int\bar f(\vec{v})\left[
      \alpha+\beta \epsilon(\vec{v})
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v +\sum_{M=1}^{K}
    \int\bar f(\vec{v})g_M(\vec{v},t)\dot g_M(\vec{v},t)\mathrm{d}^3v . \nonumber 
\end{equation}
%
On the other hand, we obtain from the restrictions the following expressions
%
\begin{eqnarray}
    \int \bar{f}(\vec{v}) g_{M}(\vec{v},t) \mathrm{d}^{3}v=\Delta_M(t) \ \  &\Rightarrow&
    \ \  \int \bar{f}(\vec{v}) \dot{g}_{M}(\vec{v},t)\mathrm{d}^{3}v=\dot{\Delta}_M(t), \nonumber \\
    \int  \bar{f}(\vec{v}) g_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v=\delta_M(t) \ \  &\Rightarrow&
    \ \  \int \bar{f}(\vec{v}) \dot{g}_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v=\dot{\delta}_M(t), \nonumber 
\end{eqnarray}
%
and as a consequence of $\sum_{M=1}^{K} \Delta_M(t) =\sum_{M=1}^{K} \delta_M(t) =0$,
we find
%
\begin{equation}
    \sum_{M=1}^{K} \dot{\Delta}_M(t)  =\sum_{M=1}^{K} \dot{\delta}_{M}(t) =0.
\end{equation}
%
Then using the previous result to (\ref{dH1-2}), we find
%
\begin{equation}
  \frac{dH'}{dt} = \sum_{M=1}^{K}\int\bar f(\vec{v})g_M(\vec{v},t)\dot g_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
Now, the summation over $M$ will be expressed in two summations
%
\begin{equation}\label{classicalH3}
    \frac{dH'}{dt}=\sum_J^{L}\int
      \bar f(\vec{v})g_J^{+}(\vec{v},t)\dot g_J^{+}(\vec{v},t)\mathrm{d}^3v
      +\sum_J^{P}\int
        \bar f(\vec{v})g_J^{-}(\vec{v},t)\dot g_J^{-}(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
where $L+P=K$. The group of $L$ cells is such that cells have an excess of
particles or/and energy (from the mean value in equilibrium). In contrast,
the group of $P$ cells is such that cells have to miss particles or/and energy
(from the mean value in equilibrium). Besides, $\dot{g}^{+}_{J}$ represents the
change on the deviation on cells that have an excess of particles or energy
while $\dot{g}^{-}_{J}$  represents the change on the deviations on cells that
have missing particles or energy. 
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells
that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents
the value of the deviation on cells that have missing particles or energy.

>>>>

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the functional \cite{bib:tolman,bib:huang}
%
\begin{equation}\label{eq:hfunctional}
   H=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Here, $f(\vec{v},t)$ is a homogeneous distribution function, \textit{i.e.}
$f$ is assumed to be independent of $\vec r$.
The functional $H$, originally introduced by Boltzmann in 1872,
describes a diluted gas, at temperature $T$, with a total energy $E$, a total
number of free classical particles $N$, and occupying a volume $V$.

@@

The distribution functions, $\{f_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. In terms of $f_M(\vec{v},t)$ and the energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    H_M' & = &  \int f_M(\vec{v},t) \ln f_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{Hcell},\\
    \mathcal{N}_M' & = & \int f_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\\
    \mathcal{E}_M' & = & \int f_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v.
\end{eqnarray}
\end{subequations}
%
Notice that assuming local homogeneous distribution functions means that we are surmising the 
local equilibrium hypothesis. In addition, $f_{M}(\vec{v},t)$ must satisfy the following
microcanonical restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

@@

We now use the variational method in order to maximize $H'$, applying the restrictions
given in Eqs.~(\ref{eq:cellrestrictions}) and (\ref{eq:micro}) through Lagrange
multipliers ($\{\alpha_M\}$ and $\{\beta_M\}$). This yields
%
\begin{eqnarray}
    \frac{\delta H'}{\delta f_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f_J(\vec{v}')}\left[
        f_M(\vec{v})\ln f_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f_M(\vec{v})}{\delta f_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f_M(\vec{v})}{\delta f_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f_J(\vec{v}') = Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') }
\end{equation}
%
where $C$ is a constant. We notice that using the variational procedure on
$H'$ predicts that the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function.

@@

We can approximate the logarithm function in (\ref{dH1-1}) to its first-order
Taylor series around $\bar f_n(\vec{v}) g_{nM}(\vec{v},t)=0$ 
%
\begin{equation}\label{lnapproximationclassical}
    \ln [\bar{f}(\vec{v})+\bar{f}(\vec{v}) g_{M}(\vec{v},t)] \approx
    \ln [\bar{f}(\vec{v})]+ g_{M}(\vec{v},t).
\end{equation}
%
With the previous approximation, (\ref{dH1-1}) obtains the following form 
%
\begin{eqnarray}
    \frac{dH'}{dt}&=&\sum_{M=1}^{K} \int \bar f(\vec{v})\left[
      1+\ln \bar f(\vec{v})+g_M(\vec{v},t)
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v,
\end{eqnarray}
%
using (\ref{eq:relacion1}) with homogeneous multipliers
%
\begin{equation}\label{dH1-2}
    \frac{dH'}{dt} = \sum_{M=1}^{K}\int\bar f(\vec{v})\left[
      \alpha+\beta \epsilon(\vec{v})
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v +\sum_{M=1}^{K}
    \int\bar f(\vec{v})g_M(\vec{v},t)\dot g_M(\vec{v},t)\mathrm{d}^3v . \nonumber 
\end{equation}
%
On the other hand, we obtain from the restrictions the following expressions
%
\begin{eqnarray}
    \int \bar{f}(\vec{v}) g_{M}(\vec{v},t) \mathrm{d}^{3}v=\Delta_M(t) \ \  &\Rightarrow&
    \ \  \int \bar{f}(\vec{v}) \dot{g}_{M}(\vec{v},t)\mathrm{d}^{3}v=\dot{\Delta}_M(t), \nonumber \\
    \int  \bar{f}(\vec{v}) g_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v=\delta_M(t) \ \  &\Rightarrow&
    \ \  \int \bar{f}(\vec{v}) \dot{g}_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v=\dot{\delta}_M(t), \nonumber 
\end{eqnarray}
%
and as a consequence of $\sum_{M=1}^{K} \Delta_M(t) =\sum_{M=1}^{K} \delta_M(t) =0$,
we find
%
\begin{equation}
    \sum_{M=1}^{K} \dot{\Delta}_M(t)  =\sum_{M=1}^{K} \dot{\delta}_{M}(t) =0.
\end{equation}
%
Then using the previous result to (\ref{dH1-2}), we find
%
\begin{equation}
  \frac{dH'}{dt} = \sum_{M=1}^{K}\int\bar f(\vec{v})g_M(\vec{v},t)\dot g_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
Now, the summation over $M$ will be expressed in two summations
%
\begin{equation}\label{classicalH3}
    \frac{dH'}{dt}=\sum_J^{L}\int
      \bar f(\vec{v})g_J^{+}(\vec{v},t)\dot g_J^{+}(\vec{v},t)\mathrm{d}^3v
      +\sum_J^{P}\int
        \bar f(\vec{v})g_J^{-}(\vec{v},t)\dot g_J^{-}(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
where $L+P=K$. The group of $L$ cells is such that cells have an excess of
particles or/and energy (from the mean value in equilibrium). In contrast,
the group of $P$ cells is such that cells have to miss particles or/and energy
(from the mean value in equilibrium). Besides, $\dot{g}^{+}_{J}$ represents the
change on the deviation on cells that have an excess of particles or energy
while $\dot{g}^{-}_{J}$  represents the change on the deviations on cells that
have missing particles or energy. 
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells
that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents
the value of the deviation on cells that have missing particles or energy.
